The Impact of Alternation

### 2017-05-15 draft.tex
ok

The model presented in this paper stems from the language inclusion
problem considered in \cite{Tacas16}. There we provide a
semi-algorithm for inclusion of data languages, based on an
exponential determinization procedure and an abstraction refinement
loop using lazy predicate abstraction \cite{blast}. In this work we
consider the full model of alternation and rely entirely on the
ability of SMT solvers to produce interpolants in the combined theory
of booleans and data. Since determinisation is not needed and
complementation is possible in linear time, the bulk of the work is
shipped to the decision procedure.

@@

A formula $\phi$ is \emph{satisfiable} in the interpretation $\I$ if
there exists a valuation $\nu$ such that $\I,\nu \models \phi$, and
\emph{valid} if $\I,\nu \models \phi$ for all valuations $\nu$.  The
\emph{theory} $\theory(\sig,\I)$ is the set of valid formulae written
in the signature $\sig$, with the interpretation $\I$. A decision
procedure for $\theory(\sig,\I)$ is an algorithm that takes a formula
$\phi$ in the signature $\sig$ and returns yes iff $\phi \in
\theory(\sig,\I)$.

@@

An \emph{alternating data automaton} (ADA or automaton in the sequel)
is a tuple $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$,
where: \begin{compactitem}
%
\item $\vec{x} \subset \vars$ is a finite set of variables of sort
  $\Data$,
%
\item $Q$ is a finite set of variables of sort $\Bool$, called
  \emph{states},
%
\item $\iota \in \Form^+(Q,\emptyset)$ is positive boolean combination
  of states defining the \emph{initial configurations},
%
\item $F \subseteq Q$ is a set of \emph{final states}, and
%
\item $\Delta : Q \times \Sigma \rightarrow
  \Form^+(Q,\overline{\vec{x}}\cup\vec{x})$ is a \emph{transition
    function},
\end{compactitem}
where $\overline{\vec{x}}=\{\overline{x} \mid x \in \vec{x}\}$. In
each formula $\Delta(q,a)$ describing a transition rule, the variables
$\overline{\vec{x}}$ track the previous and $\vec{x}$ the current
values of the variables of $\A$. Observe that the initial values of
the variables are left unconstrained, as the initial configuration
does not contain free data variables. The size of $\A$ is defined as
$\len{\A} = \len{\iota} + \sum_{(q,a) \in
  Q\times\Sigma}\len{\Delta(q,a)}$.

\begin{figure}[htb]
\centerline{\includegraphics[scale=0.8]{figures/Example.pdf}}
\caption{Alternating Data Automaton Example}
\label{fig:ex}
\end{figure}

\paragraph{\em Example}
Figure \ref{fig:ex} depicts an ADA with input alphabet $\Sigma =
\set{a,b}$, variables $\vec{x} = \set{x,y}$, states $Q =
\set{q_0,q_1,q_2,q_3,q_4}$, initial configuration $q_0$, final states
$F=\set{q_3,q_4}$ and transitions:
\[\begin{array}{rcl}
\Delta(q_0,a) & \equiv & q_1 \wedge q_2 \wedge x=0 \wedge y=0 \\
\Delta(q_1,a) & \equiv & q_1 \wedge q_3 \wedge x=\overline{y}+1 \wedge y=\overline{x}+1 \\
\Delta(q_1,b) & \equiv & q_3 \wedge \overline{x} \geq \overline{y} \\
\Delta(q_2,a) & \equiv & q_2 \wedge x > \overline{x} \wedge y > \overline{y} \\
\Delta(q_2,b) & \equiv & q_4 \wedge \overline{x}>\overline{y}
\end{array}\]
The missing rules, such as $\Delta(q_0,b)$, are assumed to be
$\bot$. Rules $\Delta(q_0,a)$ and $\Delta(q_1,a)$ are universal and
there are no existential nondeterministic rules. Rules $\Delta(q_1,a)$
and $\Delta(q_2,a)$ compare past ($\overline{x},\overline{y}$) with
present ($x,y$) values, $\Delta(q_0,a)$ constrains the present and
$\Delta(q_1,b)$, $\Delta(q_2,b)$ the past values, respectively. \qed

Formally, let $\vec{x}_k = \set{x_k \mid x \in \vec{x}}$, $k\geq0$, be
a set of time-stamped variables. For an input event $a \in \Sigma$ and
a formula $\phi$, we write $\Delta(\phi,a)$ (respectively
$\Delta^k(\phi,a)$) for the formula obtained from $\phi$ by
simultaneously replacing each state $q \in \fv{\Bool}{\phi}$ by the
formula $\Delta(q,a)$ (respectively
$\Delta(q,a)[\vec{x}_k/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}]$, for
all $k\geq0$). Given a word $w = (a_1,\nu_1)(a_2,\nu_2) \ldots
(a_n,\nu_n)$, the \emph{run} of $\A$ over $w$ is the sequence of
formulae: \[\phi_0(Q) \Rightarrow \phi_1(Q,\vec{x}_0 \cup \vec{x}_1)
\Rightarrow \ldots \Rightarrow \phi_n(Q,\vec{x}_0 \cup \ldots
\vec{x}_n)\] where $\phi_0 \equiv \iota$ and, for all $k\in[1,n]$, we
have $\phi_k \equiv \Delta^k(\phi_{k-1},a_k)$. Next, we slightly abuse
notation and write $\Delta(\iota,a_1,\ldots,a_n)$ for the formula
$\phi_n(\vec{x}_0,\ldots,\vec{x}_n)$ above. We say that $\A$
\emph{accepts} $w$ iff $\I,\nu \models \Delta(\iota,a_1,\ldots,a_n)$,
where $\nu$ is some valuation that maps:\begin{inparaenum}[(1)]
%
\item each $x \in \vec{x}_k$ to $\nu_k(x)$, for all $k\in[1,n]$, 
%
\item each $q \in \fv{\Bool}{\phi_n} \cap F$ to $\top$ and 
%
\item each $q \in \fv{\Bool}{\phi_n} \setminus F$ to $\bot$.
\end{inparaenum}
The language of $\A$ is the set $L(\A)$ of words from
$\Sigma[\vec{x}]^*$ accepted by $\A$.

\paragraph{\em Example} 
The following sequence is a non-accepting run of the ADA from Figure
\ref{fig:ex} on the word
$(a,\tuple{0,0}),(a,\tuple{1,1}),(b,\tuple{2,1})$: 
\[\begin{array}{l}
q_0 \stackrel{(a,\tuple{0,0})}{\Longrightarrow} 
q_1 \wedge q_2 \wedge x_1=0 \wedge y_1=0 \stackrel{(a,\tuple{1,1})}{\Longrightarrow} \\ 
(q_1 \wedge q_2 \wedge x_2=y_1+1 \wedge y_2=x_1+1) ~\wedge \\[-2mm]
(q_2 \wedge x_2>x_1 \wedge y_2>y_1) \wedge x_1=0 \wedge y_1=0 \stackrel{(b,\tuple{2,1})}{\Longrightarrow} \\
(q_3 \wedge x_2\geq y_2) \wedge (q_4 \wedge x_2>y_2) \wedge x_2=y_1+1 \wedge y_2=x_1+1 \\ 
\wedge~ (q_4 \wedge x_2>y_2) \wedge x_2>x_1 \wedge y_2>y_1 \wedge x_1=0 \wedge y_1=0 
\rightarrow \bot~ \hfill\text{\qed}
\end{array}\]

In this paper we tackle the following problems: \begin{compactenum}
\item \emph{boolean closure}: given automata $\A_1$ and $\A_2$, both
  with the same set of variables $\vec{x}$, do there exist automata
  $\A_\cup$, $\A_\cap$ and $\overline{\A_1}$ such that $L(\A_\cup) =
  \A_1 \cup \A_2$, $L(A_\cap) = \A_1 \cap \A_2$ and
  $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus L(\A_1)$ ?
%
\item \emph{emptiness}: given an automaton $\A$, is it the case that
  $L(\A) = \emptyset$ ?
\end{compactenum}

@@

A \emph{safety invariant} for $\A$ is a function $\Inv : (Q \mapsto
\booli) \rightarrow 2^{\vec{x} \mapsto \Data^\I}$ such that, for every
boolean valuation $\beta : Q \rightarrow \booli$, every valuation $\nu
: \vec{x} \mapsto \Data^\I$ of the data variables and every finite
sequence $u \in \Sigma^*$ of input events, the following
hold: \begin{compactenum}
%% \item $\beta \models \iota \Rightarrow \nu_0 \in \Inv(\beta)$, where
%%   $\nu_0(x) = \mathbf{0}$, for all $x \in \vec{x}$.
%
\item\label{it1:inv} $\I,\beta \cup \nu \models \Post{\A}(\iota,u)
  \Rightarrow \nu \in \Inv(\beta)$, and
%
\item\label{it2:inv} $\nu \in \Inv(\beta) \Rightarrow \I,\beta \cup
  \nu \not\models \Accept{\A}(u)$.
\end{compactenum}
If $\Inv$ satisfies only the first point above, we call it an
\emph{invariant}. Intuitively, a safety invariant maps every boolean
valuation into a set of data valuations, that contains the initial
configuration $\iota = \Post{\A}(\iota,\varepsilon)$, whose data
variables are unconstrained, over-approximates the set of reachable
valuations (point \ref{it1:inv}) and excludes the valuations satisfying the
acceptance condition (point \ref{it2:inv}).  A formula $\phi(Q,\vec{x})$
\emph{defines} $\Inv$ iff for all $\beta: Q \rightarrow \booli$ and
$\nu : \vec{x} \rightarrow \Data^\I$, we have $\I,\beta\cup\nu \models
\phi$ iff $\nu \in \Inv(\beta)$.

@@

\paragraph{\em Remark}
While computing the concrete post-image $\Post{\A}$ relies on an
expensive quantifier elimination, it is common belief that this should
hold also for $\AbsPost{\A}$. However, the choice of conjoined
predicates that form $\AbsPost{\A}$ requires only a quantifier-free
satisfiability query: $\Post{\A}(\phi,a) \models \pi \iff \exists
\overline{\vec{x}} ~.~ \Delta(\phi[\overline{\vec{x}}/\vec{x}],a)
\models \pi \iff \Delta(\phi[\overline{\vec{x}}/\vec{x}],a) \wedge
\neg\pi$ unsatisfiable. \qed

@@

\paragraph{\em Remark} 
We deviate from the standard notion of \emph{generalized interpolants}
\cite{mcmillan06}, which requires $\fv{}{I_i} \subseteq \fv{}{\phi_i}
\cap \fv{}{\phi_{i+1}}$, for each $i \in [1,n-1]$, and assume further
that the variables have been already renamed by removing the
time-stamps. \qed

@@

Given a non-empty spurious counterexample $u = a_1\ldots a_n$, where
$n>0$, we consider the following interpolation problem: 
\begin{eqnarray}\label{eq:interpolation-problem}
\Theta(u) & \equiv & \theta_0(Q_0) \wedge \theta_1(Q_0 \cup Q_1,\vec{x}_0
\cup \vec{x}_1) \wedge \ldots \\ 
&& \wedge~ \theta_n(Q_{n-1} \cup Q_n,\vec{x}_{n-1} \cup \vec{x}_n) \wedge \theta_{n+1}(Q_n) \nonumber
\end{eqnarray}
where $Q_k = \set{q_k \mid q \in Q}$, $k \in [0,n]$ are time-stamped
sets of boolean variables corresponding to the set $Q$ of states of
$\A$. The first conjunct $\theta_0(Q_0) \equiv \iota[Q_0/Q]$ is the
initial configuration of $\A$, with every $q \in \fv{\Bool}{\iota}$
replaced by $q_0$. The definition of $\theta_k$, for all $k\in[1,n]$,
uses \emph{replacement sets} $R_\ell \subseteq Q_\ell$, $\ell\in
[0,n]$, which are defined inductively below: \begin{compactitem}
\item $R_0 = \fv{\Bool}{\theta_0}$,
%
\item $\theta_\ell \equiv \bigwedge_{q_{\ell-1}\in R_{\ell-1}}
  (q_{\ell-1} \rightarrow
  \Delta(q,a_\ell)[Q_\ell/Q,\vec{x}_{\ell-1}/\overline{\vec{x}},\vec{x}_\ell/\vec{x}])$
  and $R_\ell = \fv{\Bool}{\theta_\ell} \cap Q_\ell$, for each
  $\ell\in[1,n]$.
%
\item $\theta_{n+1}(Q_n) \equiv \bigwedge_{q \in Q \setminus F} (q_n
  \rightarrow \bot)$.
\end{compactitem}
The intuition is that $R_0,\ldots,R_n$ are the sets of states replaced,
$\theta_0, \ldots, \theta_n$ are the transition rules fired on the run
of $\A$ over $u$ and $\theta_{n+1}$ is the acceptance condition, which
forces the last remaining non-final states to be false.

@@

We recall that a run of $\A$ over $u$ is a sequence: \[\phi_0(Q)
\Rightarrow \phi_1(Q,\vec{x}_0\cup\vec{x}_1) \Rightarrow \ldots
\Rightarrow \phi_n(Q,\vec{x}_0\cup\ldots\cup\vec{x}_n)\] where
$\phi_0$ is the initial configuration $\iota$ and for each $k>0$,
$\phi_k$ is obtained from $\phi_{k-1}$ by replacing each state $q \in
\fv{\Bool}{\phi_{k-1}}$ by the formula
$\Delta(q,a_k)[\vec{x}_{k-1}/\overline{\vec{x}},\vec{x}_k/\vec{x}]$,
given by the transition function of $\A$. Observe that, because the
states are replaced with transition formulae when moving one step in a
run, the formulae thus produced lose track of the control history and
are not directly suitable to producing interpolants that relate states
and data.

@@

\begin{lemma}\label{lemma:progress}
  For any non-empty sequence $u = a_1\ldots a_n \in \Sigma^*$
  such that $\Accept{\A}(u)$ is unsatisfiable, the following
  hold: \begin{compactenum}
  \item\label{it1:progress} $\Theta(u)$ is unsatisfiable, and
    %
  \item\label{it2:progress} if $\tuple{\top,I_0,\dots,I_{n},\bot}$ is
    an interpolant for $\Theta(u)$ such that $\set{I_i \mid i
      \in[0,n]} \subseteq \preds$ then $\AbsAccept{\A}(u)$ is
    unsatisfiable.
  \end{compactenum}
\end{lemma}

We have now all the ingredients to describe the first emptiness
checking semi-algorithm for alternating data
automata. Algorithm\footnote{Though termination is not guaranteed, we
  call it algorithm for conciseness.} \ref{alg:predabs} builds an
\emph{abstract reachability tree} (ART) whose nodes are labeled with
formulae over-approximating the concrete sets of configurations, and a
covering relation between nodes in order to ensure that the set of
formulae labeling the nodes in the ART forms an antichain. Any
spurious counterexample is eliminated by computing an interpolant and
adding its formulae to the set of predicates.

@@

  \If{$\Lambda(n) \wedge \bigwedge_{q \in Q \setminus F} (q
    \rightarrow \bot)$ is satisfiable} \Comment{counterexample candidate}

  \State let $u = a_1\ldots a_k$ be the label of the path from
  $\rootNode$ to $n$

  \If{$\Accept{\A}(u)$ is satisfiable} \Comment{counterexample is
    feasible}
  \label{ln:feasible}

  \State get model $\beta : Q \rightarrow \booli$, $\{\nu_i :
  \vec{x}_i \rightarrow \Data^\I \mid i=0\ldots k\}$ of $\Accept{\A}(u)$

@@

  \State $p \leftarrow \Call{FindPivot}{u,\Art}$
  \label{ln:pivot}

  \State let $v$ be the label of the path from $p$ to $n$ in $\Art$

@@

Formally, an ART is tuple $\Art =
\tuple{N,E,\rootNode,\Lambda,R,T,\lhd}$, where: \begin{compactitem}
\item $N$ is a set of nodes, 
%
\item $E \subseteq N \times \Sigma \times
N$ is a set of edges, 
%
\item $\rootNode \in N$ is the root of the directed
tree $(N,E)$, 
%
\item $\Lambda : N \rightarrow \Form(Q,\vec{x})$ is a labeling of the
  nodes with formulae, such that $\Lambda(\rootNode) = \iota$,
%
\item $R : N \rightarrow 2^Q$ is a labeling of nodes with replacement
  sets, such that $R(\rootNode) = \fv{\Bool}{\iota}$,
%
\item $T : E \rightarrow \bigcup_{i=0}^\infty
  \Form(Q_i,\vec{x}_i,Q_{i+1},\vec{x}_{i+1})$ is a labeling of edges
  with time-stamped formulae, and
%
\item $\lhd \subseteq N \times N$ is a set of
\emph{covering edges}. 
\end{compactitem}
%% For simplicity, we write $T(n,a,m)$ for $T(e)$, when $e$ is $(n,a,m)$.
%% The replacement sets and time-stamped edge labels are defined below,
%% for each edge $e = (n,a,m) \in E$, if $n$ lies at depth\footnote{We
%%   consider that the root node has zero depth.} $k$ in the tree
%% $(N,E)$ then we have:
%% \[\begin{array}{rcl}
%% T(e) & = & \bigwedge_{q \in R(n)} (q_k \rightarrow \Delta(q,a)[Q_{k+1}/Q,\vec{x}_{k}/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}]) \\
%% R(m) & = & \bigcup_{q \in R(n)}\fv{\Bool}{\Delta(q,a)}
%% \end{array}\]

>>>>

The model presented in this paper stems from the language inclusion
problem considered in \cite{Tacas16}. There we provide a
semi-algorithm for inclusion of data languages, based on an
exponential determinization procedure and an abstraction refinement
loop using lazy predicate abstraction \cite{blast}. In this work we
consider the full model of alternation and rely entirely on the
ability of SMT solvers to produce interpolants in the combined theory
of booleans and data. Since determinisation is not needed and
complementation is possible in linear time, the bulk of the work is
carried out by the decision procedure.

@@

A formula $\phi$ is \emph{satisfiable} in the interpretation $\I$ if
there exists a valuation $\nu$ such that $\I,\nu \models \phi$, and
\emph{valid} if $\I,\nu \models \phi$ for all valuations $\nu$.  The
\emph{theory} $\theory(\sig,\I)$ is the set of valid formulae written
in the signature $\sig$, with the interpretation $\I$. A \emph{decision
procedure} for $\theory(\sig,\I)$ is an algorithm that takes a formula
$\phi$ in the signature $\sig$ and returns yes iff $\phi \in
\theory(\sig,\I)$.

@@

An \emph{alternating data automaton} (ADA or automaton in the sequel)
is a tuple $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$,
where: \begin{compactitem}
%
\item $\vec{x} \subset \vars$ is a finite set of variables of sort
  $\Data$,
%
\item $Q \subset \vars$ is a finite set of variables of sort $\Bool$
  (\emph{states}),
%
\item $\iota \in \Form^+(Q,\emptyset)$ is the \emph{initial
  configuration},
%
\item $F \subseteq Q$ is a set of \emph{final states}, and
%
\item $\Delta : Q \times \Sigma \rightarrow
  \Form^+(Q,\overline{\vec{x}}\cup\vec{x})$ is a \emph{transition
    function},
\end{compactitem}
where $\overline{\vec{x}}=\{\overline{x} \mid x \in \vec{x}\}$. In
each formula $\Delta(q,a)$ describing a transition rule, the variables
$\overline{\vec{x}}$ track the previous and $\vec{x}$ the current
values of the variables of $\A$. Observe that the initial values of
the variables are left unconstrained, as the initial configuration
does not contain free data variables. The size of $\A$ is defined as
$\len{\A} = \len{\iota} + \sum_{(q,a) \in
  Q\times\Sigma}\len{\Delta(q,a)}$.

\begin{figure}[htb]
\vspace*{-\baselineskip}
\centerline{\includegraphics[scale=0.8]{figures/Example.pdf}}
\vspace*{-\baselineskip}
\caption{Alternating Data Automaton Example}
\label{fig:ex}
\end{figure}

\paragraph{\em Example}
Figure \ref{fig:ex} depicts an ADA with input alphabet $\Sigma =
\set{a,b}$, variables $\vec{x} = \set{x,y}$, states $Q =
\set{q_0,q_1,q_2,q_3,q_4}$, initial configuration $q_0$, final states
$F=\set{q_3,q_4}$ and transitions:
\[\begin{array}{rcl}
\Delta(q_0,a) & \equiv & q_1 \wedge q_2 \wedge x\teq0 \wedge y\teq0 \\
\Delta(q_1,a) & \equiv & q_1 \wedge q_3 \wedge x\teq\overline{y}+1 \wedge y\teq\overline{x}+1 \\
\Delta(q_1,b) & \equiv & q_3 \wedge \overline{x} \geq \overline{y} \\
\Delta(q_2,a) & \equiv & q_2 \wedge x > \overline{x} \wedge y > \overline{y} \\
\Delta(q_2,b) & \equiv & q_4 \wedge \overline{x}>\overline{y}
\end{array}\]
The missing rules, such as $\Delta(q_0,b)$, are assumed to be
$\bot$. Rules $\Delta(q_0,a)$ and $\Delta(q_1,a)$ are universal and
there are no existential nondeterministic rules. Rules $\Delta(q_1,a)$
and $\Delta(q_2,a)$ compare past ($\overline{x},\overline{y}$) with
present ($x,y$) values, $\Delta(q_0,a)$ constrains the present and
$\Delta(q_1,b)$, $\Delta(q_2,b)$ the past values, respectively. \qed

Formally, let $\vec{x}_k = \set{x_k \mid x \in \vec{x}}$, for any
$k\geq0$, be a set of time-stamped variables. For an input event $a
\in \Sigma$ and a formula $\phi$, we write $\Delta(\phi,a)$
(respectively $\Delta^k(\phi,a)$) for the formula obtained from $\phi$
by simultaneously replacing each state $q \in \fv{\Bool}{\phi}$ by the
formula $\Delta(q,a)$ (respectively
$\Delta(q,a)[\vec{x}_k/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}]$, for
$k\geq0$). Given a word $w = (a_1,\nu_1)(a_2,\nu_2) \ldots
(a_n,\nu_n)$, the \emph{run} of $\A$ over $w$ is the sequence of
formulae: \[\phi_0(Q) \Rightarrow \phi_1(Q,\vec{x}_0 \cup \vec{x}_1)
\Rightarrow \ldots \Rightarrow \phi_n(Q,\vec{x}_0 \cup \ldots \cup
\vec{x}_n)\] where $\phi_0 \equiv \iota$ and, for all $k\in[1,n]$, we
have $\phi_k \equiv \Delta^k(\phi_{k-1},a_k)$. Next, we slightly abuse
notation and write $\Delta(\iota,a_1,\ldots,a_n)$ for the formula
$\phi_n(\vec{x}_0,\ldots,\vec{x}_n)$ above. We say that $\A$
\emph{accepts} $w$ iff $\I,\nu \models \Delta(\iota,a_1,\ldots,a_n)$,
for some valuation $\nu$ that maps:\begin{inparaenum}[(1)]
%
\item each $x \in \vec{x}_k$ to $\nu_k(x)$, for all $k\in[1,n]$, 
%
\item each $q \in \fv{\Bool}{\phi_n} \cap F$ to $\top$ and 
%
\item each $q \in \fv{\Bool}{\phi_n} \setminus F$ to $\bot$.
\end{inparaenum}
The language of $\A$ is the set $L(\A)$ of words from
$\Sigma[\vec{x}]^*$ accepted by $\A$.

\paragraph{\em Example} 
The following sequence is a non-accepting run of the ADA from Figure
\ref{fig:ex} on the word
$(a,\tuple{0,0}),(a,\tuple{1,1}),(b,\tuple{2,1})$, where $\Data^\I =
\zed$ and the function symbols have standard arithmetic
interpretation:
\[\begin{array}{l}
q_0 \stackrel{(a,\tuple{0,0})}{\Longrightarrow} 
q_1 \wedge q_2 \wedge x_1\teq0 \wedge y_1\teq0 \stackrel{(a,\tuple{1,1})}{\Longrightarrow} \\ 
\underbrace{q_1 \wedge q_2 \wedge x_2\teq y_1+1 \wedge y_2\teq x_1+1}_{q_1} ~\wedge~
\underbrace{q_2 \wedge x_2>x_1 \wedge y_2>y_1}_{q_2} \\ 
\wedge~ x_1\teq0 \wedge y_1\teq0 \stackrel{(b,\tuple{2,1})}{\Longrightarrow} 
\underbrace{q_3 \wedge x_2\geq y_2}_{q_1} \wedge \underbrace{q_4 \wedge x_2>y_2}_{q_2} \wedge x_2\teq y_1+1 \\ 
\wedge~ y_2\teq x_1+1 \wedge \underbrace{q_4 \wedge x_2>y_2}_{q_2} \wedge x_2>x_1 \wedge y_2>y_1 \wedge x_1\teq0 \wedge y_1\teq0 ~\hfill\text{\qed}
\end{array}\]

In this paper we tackle the following problems: \begin{compactenum}
\item \emph{boolean closure}: given automata $\A_1$ and $\A_2$, both
  with the same set of variables $\vec{x}$, do there exist automata
  $\A_\cup$, $\A_\cap$ and $\overline{\A_1}$ such that $L(\A_\cup) =
  \A_1 \cup \A_2$, $L(A_\cap) = \A_1 \cap \A_2$ and
  $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus L(\A_1)$ ?
%
\item \emph{emptiness}: given an automaton $\A$, is $L(\A) =
  \emptyset$ ?
\end{compactenum}

@@

A \emph{safety invariant} for $\A$ is a function $\Inv : (Q \mapsto
\booli) \rightarrow 2^{\vec{x} \mapsto \Data^\I}$ such that, for every
boolean valuation $\beta : Q \rightarrow \booli$, every valuation $\nu
: \vec{x} \mapsto \Data^\I$ of the data variables and every finite
sequence $u \in \Sigma^*$ of input events, the following
hold: \begin{compactenum}
%% \item $\beta \models \iota \Rightarrow \nu_0 \in \Inv(\beta)$, where
%%   $\nu_0(x) = \mathbf{0}$, for all $x \in \vec{x}$.
%
\item\label{it1:inv} $\I,\beta \cup \nu \models \Post{\A}(\iota,u)
  \Rightarrow \nu \in \Inv(\beta)$, and
%
\item\label{it2:inv} $\nu \in \Inv(\beta) \Rightarrow \I,\beta \cup
  \nu \not\models \Accept{\A}(u)$.
\end{compactenum}
If $\Inv$ satisfies only the first point above, we call it an
\emph{invariant}. Intuitively, a safety invariant maps every boolean
valuation into a set of data valuations, that contains the initial
configuration $\iota \equiv \Post{\A}(\iota,\varepsilon)$, whose data
variables are unconstrained, over-approximates the set of reachable
valuations (point \ref{it1:inv}) and excludes the valuations satisfying the
acceptance condition (point \ref{it2:inv}).  A formula $\phi(Q,\vec{x})$
is said to \emph{define} $\Inv$ iff for all $\beta: Q \rightarrow \booli$ and
$\nu : \vec{x} \rightarrow \Data^\I$, we have $\I,\beta\cup\nu \models
\phi$ iff $\nu \in \Inv(\beta)$.

@@

\paragraph{\em Remark}
While computing the concrete post-image $\Post{\A}$ relies on an
expensive quantifier elimination, it is common belief that this should
hold also for $\AbsPost{\A}$. However, chosing the conjoined
predicates that form $\AbsPost{\A}$ requires only a quantifier-free
satisfiability query: $\Post{\A}(\phi,a) \models \pi$ iff $\exists
\overline{\vec{x}} ~.~ \Delta(\phi[\overline{\vec{x}}/\vec{x}],a)
\models \pi$ iff $\Delta(\phi[\overline{\vec{x}}/\vec{x}],a) \wedge
\neg\pi$ is unsatisfiable. \qed

@@

\paragraph{\em Remark} 
We deviate w.l.o.g. from the standard notion of \emph{generalized
  interpolants} \cite{mcmillan06}, which requires $\fv{}{I_i}
\subseteq \fv{}{\phi_i} \cap \fv{}{\phi_{i+1}}$, for each $i \in
          [1,n-1]$, and assume further that the variables have been
          renamed by removing the time stamps. \qed

@@

Given a non-empty spurious counterexample $u = a_1\ldots a_n$, where
$n>0$, we consider the following interpolation problem: 
\begin{eqnarray}\label{eq:interpolation-problem}
\Theta(u) & \equiv & \theta_0(Q_0) \wedge \theta_1(Q_0 \cup Q_1,\vec{x}_0
\cup \vec{x}_1) \wedge \ldots \\ 
&& \wedge~ \theta_n(Q_{n-1} \cup Q_n,\vec{x}_{n-1} \cup \vec{x}_n) \wedge \theta_{n+1}(Q_n) \nonumber
\end{eqnarray}
where $Q_k = \set{q_k \mid q \in Q}$, $k \in [0,n]$ are time-stamped
sets of boolean variables corresponding to the set $Q$ of states of
$\A$. The first conjunct $\theta_0(Q_0) \equiv \iota[Q_0/Q]$ is the
initial configuration of $\A$, with every $q \in \fv{\Bool}{\iota}$
replaced by $q_0$. The definition of $\theta_k$, for all $k\in[1,n]$,
uses \emph{replacement sets} $R_\ell \subseteq Q_\ell$, $\ell\in
[0,n]$, which are defined inductively below: \begin{compactitem}
\item $R_0 = \fv{\Bool}{\theta_0}$,
%
\item $\theta_\ell \equiv \bigwedge_{q_{\ell-1}\in R_{\ell-1}}
  (q_{\ell-1} \rightarrow
  \Delta(q,a_\ell)[Q_\ell/Q,\vec{x}_{\ell-1}/\overline{\vec{x}},\vec{x}_\ell/\vec{x}])$
  and $R_\ell = \fv{\Bool}{\theta_\ell} \cap Q_\ell$, for each
  $\ell\in[1,n]$.
%
\item $\theta_{n+1}(Q_n) \equiv \bigwedge_{q \in Q \setminus F} (q_n
  \rightarrow \bot)$.
\end{compactitem}
The intuition is that $R_0,\ldots,R_n$ are the sets of states
replaced, $\theta_0, \ldots, \theta_n$ are the sets of transition
rules fired on the run of $\A$ over $u$ and $\theta_{n+1}$ is the
acceptance condition, which forces the last remaining non-final states
to be false.

@@

We recall that a run of $\A$ over $u$ is a sequence: \[\phi_0(Q)
\Rightarrow \phi_1(Q,\vec{x}_0\cup\vec{x}_1) \Rightarrow \ldots
\Rightarrow \phi_n(Q,\vec{x}_0\cup\ldots\cup\vec{x}_n)\] where
$\phi_0$ is the initial configuration $\iota$ and for each $k>0$,
$\phi_k$ is obtained from $\phi_{k-1}$ by replacing each state $q \in
\fv{\Bool}{\phi_{k-1}}$ by the formula
$\Delta(q,a_k)[\vec{x}_{k-1}/\overline{\vec{x}},\vec{x}_k/\vec{x}]$,
given by the transition function of $\A$. Observe that, because the
states are replaced with transition formulae when moving one step in a
run, these formulae lose track of the control history and are not
suitable for producing interpolants that relate states and data.

@@

\begin{lemma}\label{lemma:progress}
  For any sequence $u = a_1\ldots a_n \in \Sigma^*$, if
  $\Accept{\A}(u)$ is unsatisfiable, the following
  hold: \begin{compactenum}
  \item\label{it1:progress} $\Theta(u)$ is unsatisfiable, and
    %
  \item\label{it2:progress} if $\tuple{\top,I_0,\dots,I_{n},\bot}$ is
    an interpolant for $\Theta(u)$ such that $\set{I_i \mid i
      \in[0,n]} \subseteq \preds$ then $\AbsAccept{\A}(u)$ is
    unsatisfiable.
  \end{compactenum}
\end{lemma}

\section{Lazy Predicate Abstraction for ADA Emptiness}

We have now all the ingredients to describe the first emptiness
checking semi-algorithm for alternating data
automata. Algorithm\footnote{Though termination is not guaranteed, we
  call it algorithm for conciseness.} \ref{alg:predabs} builds an
\emph{abstract reachability tree} (ART) whose nodes are labeled with
formulae over-approximating the concrete sets of configurations, and a
covering relation between nodes in order to ensure that the set of
formulae labeling the nodes in the ART forms an antichain. Any
spurious counterexample is eliminated by computing an interpolant and
adding its formulae to the set of predicates (cf. Lemma \ref{lemma:progress}).

@@

  \State let $\lambda(n) = a_1\ldots a_k$ be the label of the path from
  $\rootNode$ to $n$

  \If{$\AbsPost{\A}(\lambda(n))$ is satisfiable} \Comment{counterexample candidate}

  \If{$\Accept{\A}(u)$ is satisfiable} \Comment{feasible counterexample}
  \label{ln:feasible}

  \State get model $(\beta,\nu_1,\ldots,\nu_k)$ of
  $\Accept{\A}(\lambda(n))$

@@

  \State $p \leftarrow \Call{FindPivot}{\lambda(n),\Art}$
  \label{ln:pivot}

  \State $v \leftarrow \Call{LeastInfeasibleSuffix}{\lambda(n),\Art}$

@@

Formally, an ART is tuple $\Art =
\tuple{N,E,\rootNode,\Lambda,R,T,\lhd}$, where: \begin{compactitem}
\item $N$ is a set of nodes, 
%
\item $E \subseteq N \times \Sigma \times
N$ is a set of edges, 
%
\item $\rootNode \in N$ is the root of the directed
tree $(N,E)$, 
%
\item $\Lambda : N \rightarrow \Form(Q,\vec{x})$ is a labeling of the
  nodes with formulae, such that $\Lambda(\rootNode) = \iota$,
%
\item $R : N \rightarrow 2^Q$ is a labeling of nodes with replacement
  sets, such that $R(\rootNode) = \fv{\Bool}{\iota}$,
%
\item $T : E \rightarrow \bigcup_{i=0}^\infty
  \Form^+(Q_i,\vec{x}_i,Q_{i+1},\vec{x}_{i+1})$ is a labeling of edges
  with time-stamped formulae, and
%
\item $\lhd \subseteq N \times N$ is a set of
\emph{covering edges}. 
\end{compactitem}
%% For simplicity, we write $T(n,a,m)$ for $T(e)$, when $e$ is $(n,a,m)$.
%% The replacement sets and time-stamped edge labels are defined below,
%% for each edge $e = (n,a,m) \in E$, if $n$ lies at depth\footnote{We
%%   consider that the root node has zero depth.} $k$ in the tree
%% $(N,E)$ then we have:
%% \[\begin{array}{rcl}
%% T(e) & = & \bigwedge_{q \in R(n)} (q_k \rightarrow \Delta(q,a)[Q_{k+1}/Q,\vec{x}_{k}/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}]) \\
%% R(m) & = & \bigcup_{q \in R(n)}\fv{\Bool}{\Delta(q,a)}
%% \end{array}\]
